---
title: "Medage/Deathrate"
format: html
editor: visual
---

Here I will further explore my second hypothesis, that median age shares a positive correlation with the death rate of each country.

Lets set things up again:

```{r}
rm(list = ls())
library(here)
library(tidyverse)
library(ggfortify)
covid<-read.csv(here("data","covid.summary.csv"))
glimpse(covid)
```

Let's bring our plot back from before:

```{r}
ggplot(covid,aes(median_age,max.deaths/population))+
  geom_point()
```

I suspect a fairly strong positive correlation here. This would be due to elderly individuals being at higher risk. With a higher median age, it is logical to assume that there may be more elderly people in the population.

# Let's make and use our model:

We will use a linear regression again:

```{r}
medagemod<-lm(max.deaths~median_age,data = covid)
```

checking assumptions:

```{r}
autoplot(medagemod)
```

The assumptions here look great.

# Move on to the stats:

```{r}
anova(medagemod)
```

We see a significant relationship here with an F(1,183)=7.8687 and p\<.01, so we will continue after rejecting the null hypothesis that there is no relation.

```{r}
summary(medagemod)
```

We see a fairly loose correlation here, with an R2=.04. This was lower than originally anticipated, but not totally unexpected.

# A final graph to sum it up:

```{r}
ggplot(covid,aes(median_age,max.deaths/population))+
  geom_point()+
  geom_smooth(method = 'lm')+
  theme_bw()+
  ylab("Overall Covid Death Rate")+
  xlab("Median Age by Country")
```
