---
title: "Cardio/deathrate"
format: html
editor: visual
---

Here, I will further explore my first hypothesis. I claim that higher cardiovascular death rates positively affect the Covid death rate (Covid deaths are a function of cardiovascular deaths) in three different eras of Covid (early Covid, the height of Covid, and post-Covid). We'll say these eras are by the year.

The null hypothesis here is that there is no relation between cardiovascular death rates and covid death rates. The alternative hypothesis is that there is a relation.

Let's set things up:

```{r}
rm(list = ls())
library(here)
library(tidyverse)
library(ggfortify)
bigcovid<-read.csv(here("data","Ramirez-Covid-Data.csv"))
bigcovid<-bigcovid[,1:14]
bigcovid$date<-mdy(bigcovid$date)
bigcovid$location<-as.factor(bigcovid$location)
bigcovid$continent<-as.factor(bigcovid$continent)
glimpse(bigcovid)
```

Recall that this is the csv I created during my data exploration. It contains only the variables I plan to use for my hypotheses. For now, I will also need to split up the data by date. I forgot to include dates, so instead I will use an older csv for this hypothesis.

## Let's create a new data frame from this data

We will have to create groups by the date

```{r}
covid<-bigcovid|>
  group_by(location,year(date),population,cardiovasc_death_rate)|>
  summarise(covid.deaths=max(total_deaths))## here I am creating a new data frame with all the variables that I need. The year() function helps grab just the year of the dates listed
covid<-covid|>
  filter(covid.deaths>0) #I am removing any death rates equal to 0 because the chances that any location has 0 deaths is assumed to be impossible
```

Make the covid eras:

```{r}
covid$`year(date)`<-as.factor(covid$`year(date)`)
levels(covid$`year(date)`)
```

```{r}
covid$`year(date)`<-gsub("2020","Early",covid$`year(date)`)
covid$`year(date)`<-gsub("2021","Height",covid$`year(date)`)
covid$`year(date)`<-gsub("2022","Height",covid$`year(date)`)
covid$`year(date)`<-gsub("2023","Post",covid$`year(date)`)
covid$`year(date)`<-gsub("2024","Post",covid$`year(date)`)
```

Let's make a plot to start. Note that I am multiplying the covid death rate by 1000 to make it easier to understand as deaths per 1000 as opposed to deaths per person:

```{r}
ggplot(covid, aes(cardiovasc_death_rate,1000*covid.deaths/population))+
  geom_point()+
  facet_wrap(~`year(date)`)
```

There seem to be intercepts all really close to 0, but we know that may not necessarily be the case because having 0 deaths due to covid is highly unlikely. the slopes seem to be about -0.125, -0.4, and -0.75. Recall that we suppose there could be actually a negative correlation. Here, it looks like there could be a strong negative correlation in the first group, a fair negative correlation in the second, and a fair negative correlation in the third. It is important to note the cluster of points in the 2-3 hundred range on the x and less than 0.25 on the y for all three eras. If there does happen to be a relationship, differential equations could point to this range being a "sweet spot" or an equilibrium. In other words, over time, all locations could shift to this range if the correlation exists.

## Now I will further analyze these data

Creating the general linear regression models:

*Note that everything is done three times to reflect the three different time frames*

```{r}
covid.before<-covid|>
  filter(`year(date)`=="Early")
covid.height<-covid|>
  filter(`year(date)`=="Height")
covid.post<-covid|>
  filter(`year(date)`=="Post")
```

```{r}
modelearly<-glm(covid.deaths/population~cardiovasc_death_rate,data = covid.before,family = quasipoisson(link = "log"))
modelheight<-glm(covid.deaths/population~cardiovasc_death_rate,data = covid.height,family = quasipoisson(link = "log"))
modelpost<-glm(covid.deaths/population~cardiovasc_death_rate,data = covid.post,family = quasipoisson(link = "log"))
```

Checking the assumptions:

```{r}
autoplot(modelearly)
autoplot(modelheight)
autoplot(modelpost)
```

The assumptions look weird but fine. I am unsure of any better model to use and do not know how exactly to figure that out, so I will continue with the quasipoisson model

## Let's look at the stats now:

```{r}
anova(modelearly)
anova(modelheight)
anova(modelpost)
```

The relation in the early group is super significant by an F(1)=19.455 and a p value \<\<0.001. For the "height" group it is also significant, but it should be noted it is a much smaller significance with an F(1)=10.763 and p\<0.01. Finally, for the "post" group we see another very significant relation with F(1)=15.44 and p\<\<0.001. All three models allow us to reject the null hypothesis that there is no relation.

```{r}
summary(modelearly)
summary(modelheight)
summary(modelpost)
```

We now see that there are decent correlations here, with Residual Deviance values of about 0.08 0.48, and 0.51. this means that the data fit the quasipoisson distribution fairly well.

## Let's sum it up with a plot

```{r}
ggplot(covid.before,aes(cardiovasc_death_rate,1000*covid.deaths/population))+
  geom_point()+
  geom_smooth(method = 'glm')+
  theme_bw()+
  ylab("Death Rate for Early Covid")+
  xlab("Cardivascular Death Rate")
ggplot(covid.height,aes(cardiovasc_death_rate,1000*covid.deaths/population))+
  geom_point()+
  geom_smooth(method = 'glm')+
  theme_bw()+
  ylab("Death Rate for Height of Covid")+
  xlab("Cardivascular Death Rate")
ggplot(covid.post,aes(cardiovasc_death_rate,1000*covid.deaths/population))+
  geom_point()+
  geom_smooth(method = "glm")+
  theme_bw()+
  ylab("Death Rate for Post Covid")+
  xlab("Cardivascular Death Rate")
```
