---
title: "CAPDAP"
format: html
editor: visual
---

# Data and R Setup

Setting up:

```{r}
rm(list=ls())
library(tidyverse)
library(here)
library(ggfortify)
covid<-read.csv(here("data","Ramirez-Covid-Data.csv"))
```

Looking at the data:

```{r}
glimpse(covid)
```

Fixing some variables and getting rid of the weird "x" columns:

```{r}
covid<-covid[,1:14]
covid$date<-mdy(covid$date)
covid$location<-as.factor(covid$location)
covid$continent<-as.factor(covid$continent)
glimpse(covid)
```

# Now to actually look at the data

grabbing the data I want:

```{r}
covid.sum<-covid|>
  group_by(location,continent,population,cardiovasc_death_rate,median_age,human_development_index)|>
  summarise(max.cases=max(total_cases),max.deaths=max(total_deaths))
```

Save the new csv:

```{r}
write.csv(covid.sum,here("data","covid.summary.csv"),row.names = F)
```

### Exploring the stats for my hypotheses:

##### Hyp 1: Cardivasc positively affects Total deaths

```{r}
plotCV<- covid.sum|>
  filter(max.deaths<=250000)

ggplot(plotCV, aes(cardiovasc_death_rate,max.deaths/population))+
  geom_point()
```

Unclear just from looking at the plot. Maybe actually a negative correllation.

##### Hyp 2: Higher median age means higher death rate

```{r}
ggplot(covid.sum,aes(median_age,max.deaths/population))+
  geom_point()
```

I would definitely expect this to be a positive relationship.

##### Hyp 3: Higher HDI means lower cases rate

```{r}
ggplot(covid.sum,aes(human_development_index,max.cases/population))+
  geom_point()
```

Whoa! There actually seems to be a positive correlation. I imagine this could be due to countries with higher HDI's, though having higher qualities of life and health services, simply have more people who move around more. This could allow a disease like covid to spread more easily.
