---
title: "Cardio/deathrate"
format: html
editor: visual
---

Here, I will explore my first hypothesis. I claim that higher cardiovascular death rates positively affect the Covid death rate in three different eras of Covid (early Covid, the height of Covid, and post-Covid). We'll say these eras are by the year.

Let's set things up:

```{r}
rm(list = ls())
library(here)
library(tidyverse)
library(ggfortify)
bigcovid<-read.csv(here("data","Ramirez-Covid-Data.csv"))
bigcovid<-bigcovid[,1:14]
bigcovid$date<-mdy(bigcovid$date)
bigcovid$location<-as.factor(bigcovid$location)
bigcovid$continent<-as.factor(bigcovid$continent)
glimpse(bigcovid)
```

Recall that this is the csv I created during my data exploration. It contains only the variables I plan to use for my hypotheses. For now, I will also need to split up the data by date. I forgot to include dates, so instead I will use an older csv for this hypothesis.

# Let's create a new data frame from this data

We will have to create groups by the date

```{r}
covid<-bigcovid|>
  group_by(location,year(date),population,cardiovasc_death_rate)|>
  summarise(max.deaths=max(total_deaths))
```

Make the covid eras:

```{r}
covid$`year(date)`<-as.factor(covid$`year(date)`)
levels(covid$`year(date)`)
```

```{r}
covid$`year(date)`<-gsub("2020","Early",covid$`year(date)`)
covid$`year(date)`<-gsub("2021","Height",covid$`year(date)`)
covid$`year(date)`<-gsub("2022","Height",covid$`year(date)`)
covid$`year(date)`<-gsub("2023","Post",covid$`year(date)`)
covid$`year(date)`<-gsub("2024","Post",covid$`year(date)`)
```

Let's make a plot to start:

```{r}
ggplot(covid, aes(cardiovasc_death_rate,max.deaths/population))+
  geom_point()+
  facet_wrap(~`year(date)`)
```

Recall that we suppose there could be actually a negative correlation. Here, it looks like there could be a strong negative correlation in the first group, a fair negative correlation in the second, and a fair negative correlation in the third. It is important to note the cluster of points in the 2-3 hundred rang on the x and less than .00025 on the y for all three eras.

# Now I will further analyze these data

Creating the linear regression models:

*Note that everything is done three times to reflect the three different time frames* (**Question**: Is there a simpler/slicker way to do this? I imagine like a group-by() that works with lm()

```{r}
covid.before<-covid|>
  filter(`year(date)`=="Early")
covid.height<-covid|>
  filter(`year(date)`=="Height")
covid.post<-covid|>
  filter(`year(date)`=="Post")
```

```{r}
modelearly<-lm(max.deaths/population~cardiovasc_death_rate,data = covid.before)
modelheight<-lm(max.deaths/population~cardiovasc_death_rate,data = covid.height)
modelpost<-lm(max.deaths/population~cardiovasc_death_rate,data = covid.post)
```

Checking the assumptions:

```{r}
autoplot(modelearly)
autoplot(modelheight)
autoplot(modelpost)
```

For all three, the assumptions look weird, but we will continue. (**Question**:What do we do if these look *too* weird?)

# Let's look at the stats now:

```{r}
anova(modelearly)
anova(modelheight)
anova(modelpost)
```

The relation in the early group is super significant by an F(1,190)=16.326 and a p value \<\<.001. For the "height" group it is also significant, but it should be noted it is a much smaller significance with an F(1,386)=10.808 and p\<.01. Finally, for the "post" group we see another very significant relation with F=15.675 and p\<\<.001. All three models allow us to reject the null hypothesis that there is no relation.

```{r}
summary(modelearly)
summary(modelheight)
summary(modelpost)
```

We now see that there are a fairly loose negative correlations with low R2 values of .07 .02 and .04. The strongest correlation was for the early group, which was initially expected. What was also expected was the weakest correlation being the "height" group.

# Let's sum it up with a plot

```{r}
ggplot(covid.before,aes(cardiovasc_death_rate,max.deaths/population))+
  geom_point()+
  geom_smooth(method = 'lm')+
  theme_bw()+
  ylab("Death Rate for Early Covid")+
  xlab("Cardivascular Death Rate")
ggplot(covid.height,aes(cardiovasc_death_rate,max.deaths/population))+
  geom_point()+
  geom_smooth(method = 'lm')+
  theme_bw()+
  ylab("Death Rate for Height of Covid")+
  xlab("Cardivascular Death Rate")
ggplot(covid.post,aes(cardiovasc_death_rate,max.deaths/population))+
  geom_point()+
  geom_smooth(method = 'lm')+
  theme_bw()+
  ylab("Death Rate for Post Covid")+
  xlab("Cardivascular Death Rate")
```
