---
title: "Medage/Deathrate"
format: html
editor: visual
---

Here I will further explore my hypothesis that median age shares a positive correlation with the death rate of each country. So I predict that the death rate will act as a function of the median age.

The null hypothesis is that there is no correlation, and the alternative is that there is an effect of median age on the death rate.

Lets set things up again:

```{r}
rm(list = ls())
library(here)
library(tidyverse)
library(ggfortify)
covid<-read.csv(here("data","covid.summary.csv"))
glimpse(covid)
```

Let's bring our plot back from before:

```{r}
ggplot(covid,aes(median_age,1000*max.deaths/population))+
  geom_point()#I multiplied by 1000 to make this easier to understand as essentially deaths per 1000 people
```

I suspect a fairly strong positive correlation here. I see an intercept of 0, though that is not likely for any location to have 0 deaths. I also see a fairly low slope, maybe around 0.15. This would be due to elderly individuals being at higher risk. With a higher median age, it is logical to assume that there may be more elderly people in the population.

## Let's make and use our model:

I will remove any zeroes because of the above reason.

```{r}
covid<-covid|>
  filter(max.cases>0)
```

We will use a linear regression:

```{r}
medagemod<-lm(max.deaths/population~median_age, data = covid)
```

checking assumptions:

```{r}
autoplot(medagemod)
```

The assumptions here look fair, so we will keep the linear model.

## Move on to the stats:

```{r}
anova(medagemod)
```

We see a significant relationship here with an F(1,181)=161.48 and p\<\<0.001, so we will continue after rejecting the null hypothesis.

```{r}
summary(medagemod)
```

We see a fairly loose correlation here, with an R2 of about 0.47. This was lower than originally anticipated, but not totally unexpected.

## A final graph to sum it up:

```{r}
ggplot(covid,aes(median_age,1000*max.deaths/population))+
  geom_point()+
  geom_smooth(method = 'lm',fullrange=F)+
  theme_bw()+
  ylab("Covid Death Rate per T")+
  xlab("Median Age by Country")+
  ylim(0,6.5)
```
